The effects of context inference on motor learning: savings, de-adaptation and spontaneous recovery

### 2021-09-20 adaptation/article/article.tex
First modification after Stefan's comments. Small things all addressed.

In this work, we show that these unexplained phenomena, alongside savings, de-adaptation, anterograde interference and spontaneous recovery, can be explained in terms of the concept of context, an idea that has recently been gaining traction in different fields \cite[e.g.][]{Sanders_Hippocampal_2020,Hunter_Contextsensitive_2021}.

@@

When an observed outcome does not match the predictions produced by $f$, the prediction error is used to update the parameters $\phi$ of the forward model. To describe this adaptation process, our model makes uses of Bayesian inference, such that
\begin{equation}
q(\phi | \vartheta, s, c) \propto p(\vartheta | s, c, \phi)p(\phi)
\end{equation}
where $q(\phi | ...)$ represents the new estimated probability distribution over the possible values of $\phi$ after having observed the outcome $\vartheta$ of the previous motor command $c$. The likelihood $p(\vartheta | s, c, \phi)$ is the probability of observing the outcome $\vartheta$ as predicted by the forward model $f$, and $p(\phi)$ is the previous estimate of $\phi$.

@@

Context is a categorical variable $\zeta$, where the identification of a context is done through Bayesian inference:
\begin{equation}
q(\zeta | \vartheta_t, \vartheta_{t-1}, c_{t-1}) \propto p(\vartheta_t | \vartheta_{t-1}, c_{t-1})p(\zeta)
\end{equation}
where the prior distribution over contexts $p(\zeta)$ includes information from two sources: (1) an expectation of continuity, encoding the expectation that the context does not change from trial to trial, and (2) an overall expectation of observing any one context, if the context did change. Below, we discuss how these two sources of information effect different experimentally observed phenomena.

@@

\subsubsection{Action selection}
Action selection is affected by the current context via the context-dependent forward model that is used. In our model, action selection is made by building a distribution over available actions which is a weighted sum of the distributions given by the existing forward models, where the weights are obtained by the context inference component of the model. In other words, a distribution is created as follows:
\begin{equation}
p(c_t) \propto \displaystyle\sum_{\zeta \in \Phi}q(\zeta | \vartheta_t, \vartheta_{t-1}, c_{t-1}) p(c_t | \zeta) \label{eq:action-selection}
\end{equation}
where $p(c_t | \zeta)$ corresponds to \eref{eqn:forward-model}. From this distribution, the current action (motor command $c_t$) can be sampled and carried out.
\todo{Maybe add a section with the progression of a trial}
\section{Experimental results}
In this section, we present experimentally-observed phenomena in four sections, and show that context inference, as done in our model, provides a unifying explanation for all of them. In the first section, we discuss switches between contexts, and how slow context inference affects these switches. In the second section, we focus on the effects of volatility in the environment on motor adaptation. In the third section, we discuss the effects of slow context inference on action selection. Finally, in the fourth section we discuss context inference during error-clamp trials, and its effect on behavior.

@@

Our model offers a different and more parsimonious\todo{Is this the right word?} explanation for the observed results: the asymmetry is due to the existence of the baseline context. As discussed above, action selection in our model is done via a weighted average (see \eref{eq:action-selection}). In experiments without cued contexts, the baseline context O and its associated forward model $m_O$, have a non-zero probability $p(\zeta | s_t ...)$. When a new block of trials starts (e.g. in the transition from 3A to A), a switch is inferred by the model (given feedback after the first trial) and $m_O$ becomes more likely (given that $m_{3A}$ has been ruled out. Therefore, in these first trials, action selection has a component guided by the baseline model, in which no extra compensatory force is applied, effectively ``pulling'' adaptation towards zero (no compensatory force). In the first group, this initial pull towards zero accelerates the transition towards A because $3A > A > 0$, but on the second group, it slows down the switch because $A > 0 > -A$.

@@

In what follows, we dive deeper\todo{Rephrase} into the experimental literature to highlight the importance of contextual information and its effects on observed behavior during motor learning tasks, as shown by our model. In addition, we discuss several predictions that our model makes, which could be verified experimentally in future studies.

@@

\cite{Oh_Minimizing_2019} posited that the difference between their two groups (~\fref{fig:oh-2019}B and C) can be explained by considering that the smaller visual rotation of 10 degrees caused participants to update their baseline forward model instead of creating a new forward model for the context. If such an account were true, it would predict that when participants walk away from the experiment they would have trouble adapting to the no-rotation reality, having to re-learn it. Our model instead would suggest that participants would immediately switch back to their baseline movements the moment they walk away from the computer screen, as that is a clear indication that the context has switched back to baseline.\todo{@Stefan: You: It's reasonable to speak of adaptable learning rates, but their model (and, in general, the many-state linear models) have a fixed learning rate, which does not change when going back to baseline. It has been tested before they came up with the many-state models.} More generally, our model predicts that the observed difference between the two groups would disappear if clear sensory contextual cues were added to the experiment. While we believe our model provides a more general explanation for the observations by \cite{Oh_Minimizing_2019}, the effect the authors describe has been documented before in chimpanzees \citep{Kojima_Memory_2004}\todo{Maybe find more citations}; however, to our knowledge such an effect has not been observed in humans in tasks like that by \cite{Oh_Minimizing_2019}.

The paradigm by \cite{Herzfeld_memory_2014} was designed to test the hypothesis that a memory of errors affected the speed with which participants adapted to curl forces. While it has been shown that sensitivity to error can be modulated \cite{Marko_Sensitivity_2012}, and indeed in our model it depends on the amount of experience (see \sref{sec:interpreting-hyperparameters}), to our knowledge there are no other experiments suggesting that it depends on the history of errors. As before, the introduction of clear sensory contextual cues would allow us to gauge the importance of context inference and of previous errors in the observed data. The same can be said of the experiments by \cite{Davidson_Scaling_2004}, in which different contexts involve movement in different directions, which requires the use of different muscles. This alone could cause delays in switching unrelated to context inference\todo{Need citation for this, but maybe use task switching literature instead?}, but we expect the effects to be insufficient to explain the observed data.

@@

\subsubsection{Mathematical implementation}
The generalized state is given by:
\begin{equation}
z_t = \{s, \vartheta_t, \phi_t, \omega_t\}
\end{equation}
where $s$ is, e.g. in a reaching task, the current position of the hand in
Cartesian coordinates, where the origin is the starting point. We assume that
motor commands are issued every $\Delta t$ for simplicity. At the beginning of
each time interval, the context is inferred combining the contextual
information and the prediction error of the outcome of the previous motor
command. This yields:
\begin{align}
  q(\zeta_t) &= q(\zeta_t | \omega_t)q(\zeta_t | s_t, s_{t-1}, c_{t-1}) \\ \label{eqn:estimated-context}
  q(\zeta_t | \omega_t) &\propto p(\omega_t | \zeta_t)p(\zeta_t) \\
  q(\zeta_t | s_t, s_{t-1}, c_{t-1}) &\propto p(s_t | \zeta_t, s_{t-1}, c_{t-1})p(x_t)p(s_{t-1})p(\zeta_t)
\end{align}
The terms $p(\zeta_t)$ refer to the prior probability of the context at the
beginning of the trial: at the first trial, this refers to prior beliefs over
which contexts are more common/likely, which we call $p_0(\zeta_t)$. At each
subsequent trial, it incorporates the belief that has so far been accumulated,
given previous trials. Note that certain events, such as the start of a new
block of trials (see main text) can return an agent to a state in which
$p_0(\zeta_t)$ becomes relevant.

@@

The update equations for the magnitude parameters are given by:
\begin{equation}
q(\mu_f, \sigma_f | s_t, s_{t-1}, c_{t-1}) \propto p(s_t |
s_{t-1}, c_{t-1}, \zeta_t)p(\mu_f, \sigma_f) \label{eqn:context-from-x}
\end{equation}
where $p(s_t | s_{t-1}, c_{t-1}, \zeta_t)$ is a Gaussian distribution centered
around $\mu_c$ with standard deviation $\sigma_c$, which is a free parameter of
the model. It is assumed that the standard deviation over outcomes ($\sigma_c$)
is related to that of the parameter $\mu_f$ as follows:
\begin{equation}
\sigma_f = \sigma_a / \nu_f
\end{equation}
which makes $\nu_f$ a parameter to estimate. Each one of the forward models can
be updated at every trial, depending on context inference (see main text). The
updated parameters for any one forward model $i$ are given by:
\begin{align}
  \mu_\phi^{(t)} &= \frac{\nu_\phi^{(t-1)} \mu_\phi^{(t-1)} + q(\zeta_i)s_t}{\nu_\phi^{(t-1)} + q(\zeta_i)} \\
  \nu_\phi^{(t)} &= \nu_\phi^{(t-1)} + q(\zeta_i) \\
  \alpha_\phi^{(t)} &= \alpha_\phi^{(t-1)} + q(\zeta_i) / 2 \\
  \beta_\phi^{(t)} &= \beta_\phi^{(t-1)} + \frac{q(\zeta_i)\nu_\phi^{(t-1)}}{\nu_\phi^{(t-1)} +
                  q(\zeta_i)}\frac{\left(s_t - \mu_\phi^{(t-1)}\right)^2}{2}
\end{align}
where $q(\zeta_i)$ is the posterior probability of context $\zeta_i$ from
\eref{eqn:estimated-context}. Note that when $q(\zeta_i) = 0$, the model $i$ is not
updated.

\subsubsection{Interpreting the hyperparameters}
\label{sec:interpreting-hyperparameters}
\todo{Remove references to code}
Naturally, $\mu$ affects the initial estimate of the adaptation, in the same
units as the necessary adaptation. $\nu$ encodes how stable this hyperprior is:
higher values (e.g. 10,000) all but guarantee that the hypermean will not move
in the face of evidence; In principle, enough evidence should still move it,
but that will not happen during an experiment. Smaller values (e.g. 1 / force\_sd,
as is the default on the code) make the hypermean (and thus the mean) follow
evidence more freely. Note that as more observations are accumulated, $\nu$
becomes bigger and bigger, solidifying the value of the hypermean.

The hyperparameters $\alpha$ and $\beta$ have a more complex effect. Note that the
mean of a Gamma distribution is $1 / (\alpha \nu)$; this mean is being used
as the standard deviation of a Gaussian by the rest of the agent,
which makes it an important measure of uncertainty. While setting the default
hyperparameters, the values used are $\alpha = 0.5 / \sigma_0$ and
$\beta = 0.5$, where $\sigma_0$ is the \textit{a priori} estimate of the
standard deviation of the force exerted by the environment (force\_sds in the
code), which controls the initial learning rate. This makes the initial
standard deviation equal $\sigma_0$, which makes it consistent with the
fixed-force model in its interface. The 0.5 values ensure that uncertainty is
large at the beginning and during the experiment is greatly reduced, but never
to a point where it is too small. Changing this 0.5 would make the
standard deviation change more quickly, making the model more or less precise
in its predictions, independently of the volatility of the mean of the
adaptation (via the hypermean).

The baseline model defaults to different values that make it a lot more
stable. The hyperstd of the mean is set to 10,000, which makes the mean
entirely stable during the duration of the experiment. The values of $\alpha$
and $\beta$ are fixed regardless of $\sigma_0$ such that the standard deviation
is 0.001 (compare that to the size of the adaptations, around 0.0125), and the
hyperparameters of the standard deviation are stable during the experiment.

>>>>

In this work, we show that these unexplained phenomena, alongside savings, de-adaptation, anterograde interference and spontaneous recovery, can be explained in terms of the concept of context, an idea that has recently been gaining traction in different fields \cite[e.g.][]{Sanders_Hippocampal_2020,Hunter_Contextsensitive_2021}. In mechanical arm experiments, we can think of one context in which the arm exerts no force and one in which it does. By identifying the current context, a participant can know how to counteract the mechanical arm and succeed in a reaching movement. In some experiments, the context is clearly identifiable via sensory cues \cite[e.g.][]{Gandolfo_Motor_1996,Shadmehr_Adaptive_1994,Ethier_Spontaneous_2008}. In many other experiments, the context is not directly observable, and context inference takes the form of an evidence-accumulating process. In this work, we propose that this process can be observed in behavioral data and can be used to explain the phenomena outlined above.

@@

When an observed outcome does not match the predictions produced by $f$, the prediction error is used to update the parameters $\phi$ of the forward model. To describe this adaptation process, our model makes uses of Bayesian inference, such that
\begin{equation}
q(\phi | \vartheta, s, c) \propto p(\vartheta | s, c, \phi)p(\phi) \label{eqn:update-mini}
\end{equation}
where $q(\phi | ...)$ represents the new estimated probability distribution over the possible values of $\phi$ after having observed the outcome $\vartheta$ of the previous motor command $c$. The likelihood $p(\vartheta | s, c, \phi)$ is the probability of observing the outcome $\vartheta$ as predicted by the forward model $f$, and $p(\phi)$ is the previous estimate of $\phi$.

@@

Context is a categorical variable $\zeta$, where the identification of a context is done through Bayesian inference:
\begin{equation}
q(\zeta | \vartheta_t, \vartheta_{t-1}, c_{t-1}) \propto p(\vartheta_t | \vartheta_{t-1}, c_{t-1})p(\zeta)
\label{eqn:context-inference-mini}
\end{equation}
where the prior distribution over contexts $p(\zeta)$ includes information from two sources: (1) an expectation of continuity, encoding the expectation that the context does not change from trial to trial, and (2) an overall expectation of observing any one context, if the context did change. Below, we discuss how these two sources of information effect different experimentally observed phenomena.

@@

\subsubsection{Action selection}
Action selection is affected by the current context via the context-dependent forward model that is used. In our model, action selection is made by building a distribution over available actions which is a weighted sum of the distributions given by the existing forward models, where the weights are obtained by the context inference component of the model. In other words, a distribution is created as follows:
\begin{equation}
p(c_t) \propto \displaystyle\sum_{\zeta \in \Phi}q(\zeta | \vartheta_t, \vartheta_{t-1}, c_{t-1}) p(c_t | \zeta) \label{eq:action-selection}
\end{equation}
where $p(c_t | \zeta)$ corresponds to \eref{eqn:forward-model}. From this distribution, the current action (motor command $c_t$) can be sampled and carried out.

\subsection{From observation to action selection}
Each trial starts with a new observation $\vartheta_t$, which comprises contextual sensory information (when available), as well as the outcome of the previously-selected action.

The first step in the model is to perform context inference using the current observation $\vartheta_t$ according to \eref{eqn:context-inference-mini} and \eref{eqn:estimated-context}. Once a probability distribution over contexts has been created, all models are updated according to \eref{eqn:update-mini} and \eref{eqn:update-full}, based on the distribution over contexts. Finally, action selection is performed using the distribution over contexts as well as the updated forward models, following \eref{eq:action-selection}.

\section{Experimental results}
In this section, we present experimentally-observed phenomena in four sections, and show that context inference, as done in our model, provides a unifying explanation for all of them. In the first section, we discuss switches between contexts, and how slow context inference affects these switches. In the second section, we focus on the effects of volatility in the environment on motor adaptation. In the third section, we discuss the effects of slow context inference on action selection. Finally, in the fourth section we discuss context inference during error-clamp trials, and its effect on behavior.

@@

Our model offers a different explanation of the observed results, based on context inference: the asymmetry is due to the existence of the baseline context. As discussed above, action selection in our model is done via a weighted average (see \eref{eq:action-selection}). In experiments without cued contexts, the baseline context O and its associated forward model $m_O$, have a non-zero probability $p(\zeta | s_t ...)$. When a new block of trials starts (e.g. in the transition from 3A to A), a switch is inferred by the model (given feedback after the first trial) and $m_O$ becomes more likely (given that $m_{3A}$ has been ruled out. Therefore, in these first trials, action selection has a component guided by the baseline model, in which no extra compensatory force is applied, effectively ``pulling'' adaptation towards zero (no compensatory force). In the first group, this initial pull towards zero accelerates the transition towards A because $3A > A > 0$, but on the second group, it slows down the switch because $A > 0 > -A$.

@@

In what follows, we review the experimental literature to highlight the importance of contextual information and its effects on observed behavior during motor learning tasks, as shown by our model. In addition, we discuss several predictions that our model makes, which could be verified experimentally in future studies.

@@

\cite{Oh_Minimizing_2019} posited that the difference between their two groups (~\fref{fig:oh-2019}B and C) can be explained by considering that the smaller visual rotation of 10 degrees caused participants to update their baseline forward model instead of creating a new forward model for the context. If such an account were true, it would predict that when participants walk away from the experiment they would have trouble adapting to the no-rotation reality, having to re-learn it. Our model instead would suggest that participants would immediately switch back to their baseline movements the moment they walk away from the computer screen, as that is a clear indication that the context has switched back to baseline.\todo{@Stefan: It's reasonable to speak of adaptable learning rates, but their model (and, in general, the many-state linear models) have a fixed learning rate, which does not change when going back to baseline. It has been tested before they came up with the many-state models.} More generally, our model predicts that the observed difference between the two groups would disappear if clear sensory contextual cues were added to the experiment. We believe our model provides a more general and principled explanation for the observations by \cite{Oh_Minimizing_2019} and, to our knowledge, such an effect has not been observed in humans in tasks like that by \cite{Oh_Minimizing_2019}.

The paradigm by \cite{Herzfeld_memory_2014} was designed to test the hypothesis that a memory of errors affected the speed with which participants adapted to curl forces. While it has been shown that sensitivity to error can be modulated \cite{Marko_Sensitivity_2012}, and indeed in our model it depends on the amount of experience (see \sref{sec:interpreting-hyperparameters}), to our knowledge there are no other experiments suggesting that it depends on the history of errors. As before, the introduction of clear sensory contextual cues would allow us to gauge the importance of context inference and of previous errors in the observed data. The same can be said of the experiments by \cite{Davidson_Scaling_2004}.

@@

\subsubsection{Mathematical implementation}
The generalized state is given by:
\begin{equation}
z_t = \{s, \vartheta_t, \phi_t, \omega_t\}
\end{equation}
where $s$ is, e.g. in a reaching task, the current position of the hand in
Cartesian coordinates, where the origin is the starting point. We assume that
motor commands are issued every $\Delta t$ for simplicity. At the beginning of
each time interval, the context is inferred combining the contextual
information and the prediction error of the outcome of the previous motor
command. This yields:
\begin{align}
  \begin{split}
  q(\zeta_t) &= q(\zeta_t | \omega_t)q(\zeta_t | s_t, s_{t-1}, c_{t-1}) \\ \label{eqn:estimated-context}
  q(\zeta_t | \omega_t) &\propto p(\omega_t | \zeta_t)p(\zeta_t) \\
  q(\zeta_t | s_t, s_{t-1}, c_{t-1}) &\propto p(s_t | \zeta_t, s_{t-1}, c_{t-1})p(x_t)p(s_{t-1})p(\zeta_t)
  \end{split}
\end{align}
The terms $p(\zeta_t)$ refer to the prior probability of the context at the
beginning of the trial: at the first trial, this refers to prior beliefs over
which contexts are more common/likely, which we call $p_0(\zeta_t)$. At each
subsequent trial, it incorporates the belief that has so far been accumulated,
given previous trials. Note that certain events, such as the start of a new
block of trials (see main text) can return an agent to a state in which
$p_0(\zeta_t)$ becomes relevant.

@@

The update equations for the magnitude parameters are given by:
\begin{equation}
q(\mu_f, \sigma_f | s_t, s_{t-1}, c_{t-1}) \propto p(s_t |
s_{t-1}, c_{t-1}, \zeta_t)p(\mu_f, \sigma_f) \label{eqn:context-from-x}
\end{equation}
where $p(s_t | s_{t-1}, c_{t-1}, \zeta_t)$ is a Gaussian distribution centered
around $\mu_c$ with standard deviation $\sigma_c$, which is a free parameter of
the model. It is assumed that the standard deviation over outcomes ($\sigma_c$)
is related to that of the parameter $\mu_f$ as follows:
\begin{equation}
\sigma_f = \sigma_a / \nu_f
\end{equation}
which makes $\nu_f$ a parameter to estimate. Each one of the forward models can
be updated at every trial, depending on context inference (see main text). The
updated parameters for any one forward model $i$ are given by:
\begin{align}
  \begin{split}
  \mu_\phi^{(t)} &= \frac{\nu_\phi^{(t-1)} \mu_\phi^{(t-1)} + q(\zeta_i)s_t}{\nu_\phi^{(t-1)} + q(\zeta_i)} \\
  \nu_\phi^{(t)} &= \nu_\phi^{(t-1)} + q(\zeta_i) \\
  \alpha_\phi^{(t)} &= \alpha_\phi^{(t-1)} + q(\zeta_i) / 2 \\
  \beta_\phi^{(t)} &= \beta_\phi^{(t-1)} + \frac{q(\zeta_i)\nu_\phi^{(t-1)}}{\nu_\phi^{(t-1)} +
    q(\zeta_i)}\frac{\left(s_t - \mu_\phi^{(t-1)}\right)^2}{2}  \label{eqn:update-full}
  \end{split}
\end{align}
where $q(\zeta_i)$ is the posterior probability of context $\zeta_i$ from
\eref{eqn:estimated-context}. Note that when $q(\zeta_i) = 0$, the model $i$ is not
updated.

\subsubsection{Interpreting the hyperparameters}
\label{sec:interpreting-hyperparameters}
$\mu$ determines the initial estimate of the adaptation, in the same units as
the necessary adaptation. $\nu$ encodes how stable this hyperprior is: higher
values (e.g. 10,000) all but guarantee that the hypermean will not change its
value after observations; In principle, enough evidence should still modify it,
but that would not happen during an experiment. Smaller values (i.e. $\sim 1$)
make the hypermean follow evidence more freely. Note that as more observations
are accumulated, $\nu$ becomes bigger and bigger, stabilizing the value of the
hypermean.

The hyperparameters $\alpha$ and $\beta$ have a more complex effect. Note that
the mean of a gamma distribution is $1 / (\alpha \nu)$; this mean is being used
as the standard deviation of a Gaussian by the rest of the agent, which makes it
an important measure of uncertainty. While setting the default hyperparameters,
the values used are $\alpha = 0.5 / \sigma_0$ and $\beta = 0.5$, where
$\sigma_0$ is the \textit{a priori} estimate of the standard deviation of the
force exerted by the environment, which controls the initial learning rate. This
makes the initial standard deviation equal $\sigma_0$, which makes it consistent
with the fixed-force model. The 0.5 values ensure that uncertainty is large at
the beginning and is greatly reduced during the experiment, but never to a point
where it is so small that it makes trial-to-trial variation in the environment
surprising. Changing this 0.5 would make the standard deviation change more
quickly, making the model more or less precise in its predictions, independently
of the volatility of the mean of the adaptation (via the hypermean).

The baseline model defaults to different values that make it a lot more
stable. The hyper-standard deviation of the mean is set to 10,000, which makes
the mean entirely stable during the duration of the experiment. The values of
$\alpha$ and $\beta$ are fixed regardless of $\sigma_0$ such that the standard
deviation is 0.001 (compared that to the size of the adaptations in mechanical
arm experiments, around 0.0125), and the hyperparameters of the standard
deviation are stable during the experiment.
